Trygve's Problem

### 2018-03-23 generativemodelsforscrna/outline.tex
Added cell indexing

Towards this end, we propose the following generative model for gene expression of a single cell.  Let $H$ denote a tree of cell types.

\begin{enumerate}
\item $T \sim \Categorical(\pi)$, one of the leaves of $H$.  Let $t_0,t_1(T),t_2(T),\cdots ,t_\ell(T)$, indicate the nodes in the unique path in $H$ which leads from the root to the leaf $T$.  Here $t_0$ is the root of the tree, $\ell$ is the depth of the leaf $T$, and $t_\ell(T)=T$.   To consolidate notation, we take $T_i \triangleq t_i(T)$.
\item For each $i<\ell$ let $Z_{T_i} \sim \cN(\mu_{T_{i+1}},\Sigma_{T_{i+1}})$, and take $Z_{T_\ell} \sim \cN(0,I)$
\item For each gene $g$, let $\lambda_g \in \mathbb R^{p}$ denote a set of parameters.  These will be a deterministic function of the $Z$s:
\[
\lambda_g = f_g \left(\sum_{i=0}^{\ell} h_{T_i}Z_{T_i}\right)
\]
\item Conditioned on $Z$, the  $g$th gene of the cell be distributed according to some $p(x_g ; \lambda_g)$. 
\end{enumerate}

@@

\begin{itemize}
\item The latent representation $Z_i$ corresponds to features of that cell which are specific to all the cells which are descendents of $T_i$.
\item The the cell's subcluster $T_{i+1}$ effects the distribution of its latent features at the $i$th level, $Z_i$.  That is, the latent distribution on the $i$th latent representation is governed by what subcluster it is part of within $T_i$.
\item The distribution of the final latent representation, $Z_T$, is simply a standard normal.
\item The functions $h_{T_i}$ indicates how these cell-type-specific features alter the distribution on the gene expressions for cells in that branch of the tree.  
\end{itemize}

@@

At the current moment, we have only actually implemented this model in the case the $H$ is a tree with depth 1, i.e. the hierarchy is trivial.   We take the data model to be a zero-inflated negative binomial.  It looks fairly promising.  For a simple qualitative assessment, consider the following.

>>>>

Towards this end, we propose the following generative model for gene expression of a single cell.  We will denote the expression of gene $g$ in cell $c$ by $X_g^c$.  Let $H$ denote a tree of cell types.

\begin{enumerate}
\item $T^c \sim \Categorical(\pi)$, one of the leaves of $H$.  Let $t_0,t_1(T^c),t_2(T^c),\cdots ,t_\ell^c(T^c)$, indicate the nodes in the unique path in $H$ which leads from the root to the leaf $T$.  Here $t_0$ is the root of the tree, $\ell^c$ is the depth of the leaf $T^c$, and $t^c_\ell(T^c)=T^c$.   To consolidate notation, we take $T^c_i \triangleq t_i(T^c)$.
\item For each $i<\ell$ let $Z^c_{T^c_i} \sim \cN(\mu_{T^c_{i+1}},\Sigma_{T^c_{i+1}})$, and take $Z^c_{T^c_\ell} \sim \cN(0,I)$
\item For each gene $g$, let $\lambda^c_g \in \mathbb R^{p}$ denote a set of parameters.  These will be a deterministic function of the $Z$s:
\[
\lambda^c_g = f_g \left(\sum_{i=0}^{\ell} h_{T^c_i}(Z^c_{T^c_i})\right)
\]
\item Conditioned on $Z^c$ we will have $X^c_g \sim p(x_g ; \lambda^c_g)$. 
\end{enumerate}

@@

\begin{itemize}
\item The latent representation $Z^c_i$ corresponds to features of that cell which are specific to all the cells which are descendents of $T^c_i$.
\item The the cell's subcluster $T^c_{i+1}$ effects the distribution of its latent features at the $i$th level, $Z^c_i$.  That is, the latent distribution on the $i$th latent representation is governed by what subcluster it is part of within $T^c_i$.
\item The distribution of the final latent representation, $Z^c_T$, is simply a standard normal.
\item The functions $h_{T^c_i}$ indicates how these cell-type-specific features alter the distribution on the gene expressions for cells in that branch of the tree.  
\end{itemize}

@@

At the current moment, we have only actually implemented this model in the case the $H$ is a tree with depth 2 (i.e. one root node and some children).  We take the data model to be a zero-inflated negative binomial.  It looks fairly promising.  For a simple qualitative assessment, consider the following.

### 2018-03-23 trygve/outline.tex
Added cell indexing

There is one possibly confusing aspect of our output that might be nice to address.  The upper and lower bounds (as we have visualized them) are each actually impossible as values for $q$.  For example, the rows of the upper bounds generally sum to more than 1.  This is because we have taken \emph{elementwise bounds} on each entry.  However, it would be nice to sample uniformly from the set of possible values of $q$ which are consistent with our observations.  MCMC or rejection sampling might be good options to investigate.  These would help give the user another way to understand what possible values $q$ might take on.

>>>>

There is one possibly confusing aspect of our output that we would like to to address.  The upper and lower bounds (as we have visualized them) are each actually impossible as values for $q$.  For example, the rows of the upper bounds generally sum to more than 1.  This is because we have taken \emph{elementwise bounds} on each entry.  An alternative is to  sample uniformly from the set of possible values of $q$ which are consistent with our observations.  We hope this may give the user another way to understand what possible values $q$ might take on.
